---
title: "ERU Project - data cleaning & exploratory analysis"
author: "Joy Wu"
date: "16 August 2015"
output: html_document
---

```{r, echo=FALSE, message = FALSE, warning = FALSE}
setwd("~/Documents/ERU project")
library(survival)
library(xtable)
library(Hmisc)
library(gridBase, gridExtra)
library(lattice)
library(lubridate)
library(dplyr)
```

# Introduction:
ERU stands for Extended Recovery Unit; a '23-hour' unit. This is a new admission category for patients who need a slightly longer overnight monitoring period post-operative but do not necessarily need to be admitted for over 24 hours. However, since the introduction of ERU admissions, the rate of actual 'on-time' discharges have been low.

# Objectives of this study are:
1. To find out the various reasons why ERU patients were discharged late or admitted inpatient on post-operative day one.
2. Come up with a pre-operative prediction model to better assess whether a patient is likely able to be discharged on time.

# Method:
- Data source: Partner's LMR database and nursing log on ERU
- Pre-op co-morbidities, various functional scores in pre-op nursing assessment form, OR service and payor information, summary of progress notes throughout PACU and ERU stay, as well as outpatient notes available on LMR are reviewed.
- The main outcome is whether the patient is discharged on time, late or admitted inpatient
- Secondary outcomes include:
1. late minute foley placement on POD1
2. Re-admission or ED visit within 3 days of discharge
3. Patient stating not ready for discharge when phone followed up at 2 days.

# Pre-lim Results:
- Patients who were listed down as ERU admission pre-op from 5/5/15 to 7/16/14 are included in this study.
- 16 Patients were excluded due to no LMR notes uploaded
- 4 Patients were excluded because of invalid MRN
- 1 patient was excluded because his/her operation was postponed after pre-op visit and his/her admission category was changed to inpatient post-op
- 4 records were duplicates so are also not included
- A total of 281 patients are included in the following analysis.

# Data cleaning and rough analysis:
The analysis is done in R. Aim to do some logistic regression eventually. Currently still waiting to clean up co-morbidity data first but here's what we have so far:

```{r, echo =FALSE, message = FALSE, warning = FALSE}
# Preparing data for analysis
# Reading in dataset
eru <- read.csv("ERU_sorted_data.csv", sep = ",")
#str(eru)

# De-identify dataset
eru$ID <- 1:nrow(eru)
eru <- select(eru, ID, ERU_outcome:education, blind_HOH:patient_not_ready4dc)
```

## Overral structure of the raw ERU dataset:
```{r}
str(eru)
```

```{r, echo =FALSE, message = FALSE, warning = FALSE, results='hide'}
# Renaming levels of ERU_outcome to something more descriptive
levels(eru$ERU_outcome)
levels(eru$ERU_outcome)[2] <- "Late"
levels(eru$ERU_outcome)[1] <- "On_time"
```

## Rough summary of ERU outcome: On_time, Late, or IP (admitted inpatient)
```{r}
# Number of patients in each category
summary(eru$ERU_outcome)

# Proportion of patients in each category
prop.table(table(eru$ERU_outcome))
paste(round(prop.table(table(eru$ERU_outcome))*100,2), "%")
```

```{r, , echo =FALSE, message = FALSE, warning = FALSE}
# Seperating ueful data chunks: -- all moved to later sections

# demographics:
# demo <- select(eru, ID:education, smoker)
# Funcational baseline:
# fnbase <- select(eru, ERU_outcome, blind_HOH:ASA_34)
# OR related information:
# ORinfo <- select(eru, ID:ERU_outcome, service:OR_date)
# Post-op timeline
# postop_chron <- select(eru, ID:ERU_outcome, TimeoutOR_after1800, ERU_after2100, PACU_event:MDplan_after7am)
# Reasons for late discharge
# reasons <- select(eru, ID:ERU_outcome, Prim_reason, other_reasons, doc_ride_issue, last_minute_foley)
# ERU discharge quality indicators
# qualty_dc <- select(eru, ID:ERU_outcome, last_minute_foley:patient_not_ready4dc)
```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Demographics, education, employment data cleaning
# demographics:
demo <- select(eru, ID:education, smoker)
# Summarize demographics
head(demo)
str(demo)

# Clean age
summary(demo$Age)
summary(is.na(demo$Age))

# Clean genders
levels(demo$Gender)
levels(demo$Gender)[2] <- "F"
summary(demo$Gender)
demo$Gender <- as.character(demo$Gender)
demo$Female <- NULL
for (i in 1:nrow(demo)) {
  if (demo$Gender[i] == "F") {
    demo$Female[i] <- 1
  }
  else {
    demo$Female[i] <- 0
  } 
}  
sum(demo$Female, na.rm = T)
summary(demo$Female)
paste(round(mean(demo$Female, na.rm = T)*100, 2), "%", sep = "")

# Clean marital status
levels(demo$Married)
levels(demo$Married)[3] <- NA
demo$Married <- as.character(demo$Married)
demo$Married <- as.integer(demo$Married)
sum(demo$Married, na.rm = T)
summary(demo$Married)
paste(round(mean(demo$Married, na.rm = T)*100, 2), "%", sep = "")

# Clean language
which(is.na(demo$no_English))
sum(demo$no_English, na.rm = T)
summary(demo$no_English)
paste(round(mean(demo$no_English, na.rm = T)*100, 2), "%", sep = "")
# No missing values in language
#demo$no_English <- as.factor(demo$no_English)

# Clean smoking status
levels(demo$smoker)
levels(demo$smoker)[3] <- NA
demo$smoker <- as.character(demo$smoker)
demo$smoker <- as.integer(demo$smoker)
sum(demo$smoker, na.rm = T)
summary(demo$smoker)
paste(round(mean(demo$smoker, na.rm = T)*100, 2), "%", sep = "")

# Simply race data 
levels(demo$Race)
summary(demo$Race)
# To White, Hispanic, Black, Asian, Other, NA
levels(demo$Race)[9] <- "Other" #level 9: White_Native American
# levels(demo$Race)[7] <- NA    #level 7: NULL (ie unknown)
levels(demo$Race)[c(5, 6)] <- "Other" #levels 5, 6: Middle Eastern, Native American
levels(demo$Race)[c(1, 2)] <- "Other"  #levels 1, 2: Arab, Asian
summary(demo$Race)
paste(round(prop.table(table(demo$Race))*100, 2), "%", sep = "")
# Tease out race data
demo$Race <- as.character(demo$Race)
demo$White <- NULL
demo$Hispanic <- NULL
demo$Black <- NULL
demo$Other <- NULL
demo$Missing_race <- NULL
for (i in 1:nrow(demo)) {
if (demo$Race[i] == "White") {
    demo$White[i] <- 1
  }
  else {
    demo$White[i] <- 0
  } 
if (demo$Race[i] == "Black") {
    demo$Black[i] <- 1
  }
  else {
    demo$Black[i] <- 0
  }
if (demo$Race[i] == "Hispanic") {
    demo$Hispanic[i] <- 1
  }
  else {
    demo$Hispanic[i] <- 0
  }
if (demo$Race[i] == "Other") {
    demo$Other[i] <- 1
  }  
  else {
    demo$Other[i] <- 0
  }  
if (demo$Race[i] == "NULL") {
    demo$Missing_race[i] <- 1
  }  
  else {
    demo$Missing_race[i] <- 0
  }
}

missing_race <- demo$Race == "NULL"
demo$White[missing_race] <- NA
demo$Black[missing_race] <- NA
demo$Hispanic[missing_race] <- NA
demo$Other[missing_race] <- NA

# For employment_status, 0 = employed/student, 1 = retired, 2 = unemployed/disabled/homemaker
# demo$employment_status <- as.factor(demo$employment_status)
levels(demo$employment_status)
levels(demo$employment_status)[c(3)] <- "unemployed/disabled"
levels(demo$employment_status)[c(2)] <- "retired"
levels(demo$employment_status)[c(1)] <- "employed"
summary(demo$employment_status)
paste(round(prop.table(table(demo$employment_status))*100,2),"%")
demo$employment_status <- as.character(demo$employment_status)
demo$employed <- NULL
demo$retired <- NULL
demo$unemployed <- NULL
demo$missing_employment <- NULL
for (i in 1:nrow(demo)) {
  if (demo$employment_status[i] == "employed") {
    demo$employed[i] <- 1
  }
  else {
    demo$employed[i] <- 0
  } 
  if (demo$employment_status[i] == "retired") {
    demo$retired[i] <- 1
  }
  else {
    demo$retired[i] <- 0
  } 
  if (demo$employment_status[i] == "unemployed/disabled") {
    demo$unemployed[i] <- 1
  }
  else {
    demo$unemployed[i] <- 0
  } 
  if (demo$employment_status[i] == "NULL") {
    demo$missing_employment[i] <- 1
  }
  else {
    demo$missing_employment[i] <- 0
  } 
} 
missing_empl <- demo$employment_status == "NULL"
demo$employed[missing_empl] <- NA
demo$retired[missing_empl] <- NA
demo$unemployed[missing_empl] <- NA



# For education, 0 = beyond high school, 1 = up to high school, 2 = 8th grade or less
# demo$education <- as.factor(demo$education)
levels(demo$education)
levels(demo$education)[c(3)] <- "grade8_or_less"
levels(demo$education)[c(2)] <- "up_to_high_school"
levels(demo$education)[c(1)] <- "beyond_high_school"
summary(demo$education)
paste(round(prop.table(table(demo$education))*100,2),"%")
demo$education <- as.character(demo$education)
demo$beyond_high_school <- NULL
demo$up_to_high_school <- NULL
demo$grade8_or_less <- NULL
demo$missing_edu <- NULL
for (i in 1:nrow(demo)) {
  if (demo$education[i] == "grade8_or_less") {
    demo$grade8_or_less[i] <- 1
  }
  else {
    demo$grade8_or_less[i] <- 0
  } 
  if (demo$education[i] == "up_to_high_school") {
    demo$up_to_high_school[i] <- 1
  }
  else {
    demo$up_to_high_school[i] <- 0
  } 
  if (demo$education[i] == "beyond_high_school") {
    demo$beyond_high_school[i] <- 1
  }
  else {
    demo$beyond_high_school[i] <- 0
  }   
  if (demo$education[i] == "NULL") {
    demo$missing_edu[i] <- 1
  }
  else {
    demo$missing_edu[i] <- 0
  }
}
missing_edu <- demo$education == "NULL"
demo$grade8_or_less[missing_edu] <- NA
demo$up_to_high_school[missing_edu] <- NA
demo$beyond_high_school[missing_edu] <- NA

# Summarise demo
str(demo)
clean.demo <- select(demo, ERU_outcome, Age, Female, Married, smoker, no_English, White:missing_edu)  

# Group by ERU outcome
by_ERUoutcome <- group_by(clean.demo, ERU_outcome)
# Useful functions: (Somehow - dplyr doesn't seem to work within functions well)
proptest_var <- function(var){
var_sum = paste(sum(var, na.rm = T), " (", round(mean(var, na.rm = T)*100,2), "%)", " [", round(prop.test(sum(var, na.rm = T), length(var), p = mean(var, na.rm = T), correct = T)$conf.int[1]*100, 2), "%-", round(prop.test(sum(var, na.rm = T), length(var), p = mean(var, na.rm = T), correct = T)$conf.int[2]*100, 2), "%]", sep = "")
}
ttest_var <- function(var){
var_sum = paste(round(mean(var, na.rm = T),2), " [", round(t.test(var, na.rm = T)$conf.int[1], 2), "-", round(t.test(var, na.rm = T)$conf.int[2], 2), "%]", sep = "")
}

# Tables:

# Demographics:
demo.brkdown <- as.data.frame(summarise(by_ERUoutcome,
    Age = ttest_var(Age),
    Female= proptest_var(Female),
    Married = proptest_var(Married),
    Smoker = proptest_var(smoker),
    No_English = proptest_var(no_English)
    ))
demo.overall <- data.frame(summarise(clean.demo,
    ERU_outcome = "Overall",                                 
    Age = ttest_var(Age),
    Female= proptest_var(Female),
    Married = proptest_var(Married),
    Smoker = proptest_var(smoker),
    No_English = proptest_var(no_English) 
  ))
Missing_demo_info <- as.data.frame(t(summarise(by_ERUoutcome,
      Age = paste(round(length(which(is.na(Age)))/length(Age)*100, 2), "%"),
      Gender = paste(round(length(which(is.na(Female)))/length(Female)*100, 2), "%"),
      Marital_status = paste(round(length(which(is.na(Married)))/length(Married)*100, 2), "%"),
      Smoking_status = paste(round(length(which(is.na(smoker)))/length(smoker)*100, 2), "%"),
      no_English = paste(round(length(which(is.na(no_English)))/length(no_English)*100, 2), "%"),
      Race = paste(round(length(which(is.na(White)))/length(White)*100, 2), "%"),
      Education = paste(round(length(which(is.na(grade8_or_less)))/length(grade8_or_less)*100, 2), "%"),
      Employment = paste(round(length(which(is.na(employed)))/length(employed)*100, 2), "%")
)))

# Race distributions:
race.brkdown <- as.data.frame(summarise(by_ERUoutcome,
    White = proptest_var(White),
    Black = proptest_var(Black),
    Hispanic = proptest_var(Hispanic),
    Other = proptest_var(Other)
))
race.overall <- as.data.frame(summarise(clean.demo,
    ERU_outcome = "Overall",                                      
    White = proptest_var(White),
    Black = proptest_var(Black),
    Hispanic = proptest_var(Hispanic),
    Other = proptest_var(Other)
))
#Run output with "options(dplyr.width = Inf)" to get full table

# Education:
edu.brkdown <- as.data.frame(summarise(by_ERUoutcome,
    Grade8_or_less = proptest_var(grade8_or_less),
    High_school = proptest_var(up_to_high_school),
    College_plus = proptest_var(beyond_high_school)
    ))
edu.overall <- as.data.frame(summarise(clean.demo,
    ERU_outcome = "Overall", 
    Grade8_or_less = proptest_var(grade8_or_less),
    High_school = proptest_var(up_to_high_school),
    College_plus = proptest_var(beyond_high_school)
    ))

# Employment:
employ.brkdown <- as.data.frame(summarise(by_ERUoutcome,
    Employed = proptest_var(employed),
    Retired = proptest_var(retired),
    Unemployed = proptest_var(unemployed)
    ))
employ.overall <- as.data.frame(summarise(clean.demo,
    ERU_outcome = "Overall", 
    Employed = proptest_var(employed),
    Retired = proptest_var(retired),
    Unemployed = proptest_var(unemployed)
    ))

```

## Demographics, education, employment split by ERU outcome
```{r, results='markup'}
# Age, gender, marital status, smoking status and langauge
demographics <- data.frame(t(demo.brkdown), t(demo.overall)); demographics

# Race and missing information
Race <- data.frame(t(race.brkdown), t(race.overall)); Race

# Education and missing information 
education <- data.frame(t(edu.brkdown), t(edu.overall)); education

# Employment status and missing information
employment <- data.frame(t(employ.brkdown), t(employ.overall)); employment

# Missing information
Missing_demo_info

```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Data cleaning
# Funcational baseline:
fnbase <- select(eru, ERU_outcome, blind_HOH:ASA_34)
str(fnbase)

# No missing values in blind_HOH
which(is.na(fnbase$blind_HOH))

# alone_postop has missing values
levels(fnbase$alone_postop)
levels(fnbase$alone_postop)[3] <- NA
fnbase$alone_postop <- as.character(fnbase$alone_postop)
fnbase$alone_postop <- as.integer(fnbase$alone_postop)

# chronic_pain currently also a factor with "NULL" or missing values
levels(fnbase$chronic_pain)
levels(fnbase$chronic_pain)[3] <- NA
fnbase$chronic_pain <- as.character(fnbase$chronic_pain)
fnbase$chronic_pain <- as.integer(fnbase$chronic_pain)

# Morse Fall risk probably should be in numeric or integer
levels(fnbase$MorseFall_risk)
levels(fnbase$MorseFall_risk)[17] <- NA
levels(fnbase$MorseFall_risk)[9:16] <- 1
levels(fnbase$MorseFall_risk)[1:8] <- 0
fnbase$MorseFall_risk <- as.character(fnbase$MorseFall_risk)
fnbase$MorseFall_risk <- as.integer(fnbase$MorseFall_risk)
names(fnbase)[names(fnbase)=="MorseFall_risk"] <- "MorseFallrisk_45plus"

# Same with mets score, BMI and ASA
levels(fnbase$METS_less4)[3] <- NA
fnbase$METS_less4 <- as.character(fnbase$METS_less4)
fnbase$METS_less4 <- as.integer(fnbase$METS_less4)
levels(fnbase$BMI_over30)[3] <- NA
fnbase$BMI_over30 <- as.character(fnbase$BMI_over30)
fnbase$BMI_over30 <- as.integer(fnbase$BMI_over30)
levels(fnbase$ASA_34)[3] <- NA
fnbase$ASA_34 <- as.character(fnbase$ASA_34)
fnbase$ASA_34 <- as.integer(fnbase$ASA_34)

df <- fnbase[, 2:10]
func_baseline_overall <- do.call(data.frame, list(
    proportion = paste(round(apply(df, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(df, 2, length),
    n = apply(df, 2, sum, na.rm = TRUE),
    missing_value = apply (df, 2, 
    function(x) sum(length(which(is.na(x))))
                )))

df_ontime <- fnbase[fnbase$ERU_outcome == "On_time", 2:10]
func_ontime <- do.call(data.frame, list(
    proportion = paste(round(apply(df_ontime, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(df_ontime, 2, length),
    n = apply(df_ontime, 2, sum, na.rm = TRUE),
    missing_value = apply (df_ontime, 2, 
    function(x) sum(length(which(is.na(x))))
                )))
df_late <- fnbase[fnbase$ERU_outcome == "Late", 2:10]
func_late <- do.call(data.frame, list(
    proportion = paste(round(apply(df_late, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(df_late, 2, length),
    n = apply(df_late, 2, sum, na.rm = TRUE),
    missing_value = apply (df_late, 2, 
    function(x) sum(length(which(is.na(x))))
                )))
df_IP <- fnbase[fnbase$ERU_outcome == "IP", 2:10]
func_IP <- do.call(data.frame, list(
    proportion = paste(round(apply(df_IP, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(df_IP, 2, length),
    n = apply(df_IP, 2, sum, na.rm = TRUE),
    missing_value = apply (df_IP, 2, 
    function(x) sum(length(which(is.na(x))))
                )))

```

## Functional baseline split by ERU outcome
```{r, results='markup'}
# Overall functional baseline of all patients
func_baseline_overall

# Functional baseline of patients who went home on time
func_ontime

# Functional baseline of patients who went home late
func_late

# Functional baseline of patients who were admitted inpatient
func_IP
```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Data cleaning
# OR related information:
ORinfo <- select(eru, ID:ERU_outcome, service:OR_date)
str(ORinfo)
levels(ORinfo$payer)[14] <- "UNKNOWN"
levels(ORinfo$payer)[10] <- "MCARE"
levels(ORinfo$payer)[c(3, 4)] <- "BLUE CROSS"
# Changing OR date info to day of week
ORinfo$OR_date <- as.character(ORinfo$OR_date)
ORinfo$OR_date <- as.Date(ORinfo$OR_date, "%m/%d/%Y")
ORinfo$weekday <- wday(ORinfo$OR_date, label = TRUE)

# Split by ERU_category
days_ontime <- ORinfo[ORinfo$ERU_outcome == "On_time", ]
days_late <- ORinfo[ORinfo$ERU_outcome == "Late", ]
days_IP <- ORinfo[ORinfo$ERU_outcome == "IP", ]
```

## OR service, payer and day of week:
```{r, results='markup'}
# Service barplot overall
par(las=2)
par(mar=c(5,4,4,2))
barplot(table(ORinfo$service), horiz = TRUE, cex.names=0.8)
title(main = "Overrall ERU patients by service", xlab = "Number of patients")

# Service barplot: Late patients
barplot(table(days_late$service), horiz = TRUE, cex.names=0.8)
title(main = "ERU patients discharged late - by service", xlab = "Number of patients")

# Table of late and admitted patients by service
service <- NULL
n_total <- NULL
n_late <- NULL
percent_late <- NULL
n_admitedIP <- NULL
percent_IP <- NULL
for(i in 1:length(levels(ORinfo$service))){
  service[i] <- levels(ORinfo$service)[i]
  n_total[i] <- nrow(ORinfo[ORinfo$service == levels(ORinfo$service)[i],])
  n_late[i] <- nrow(ORinfo[ORinfo$service == levels(ORinfo$service)[i] & ORinfo$ERU_outcome == "Late",])
  percent_late[i] <- paste(round(n_late[i]/n_total[i]*100, 0), "%")
  n_admitedIP[i] <- nrow(ORinfo[ORinfo$service == levels(ORinfo$service)[i] & ORinfo$ERU_outcome == "IP",])
  percent_IP[i] <- paste(round(n_admitedIP[i]/n_total[i]*100, 0), "%")
}
late_prop_service <- data.frame(service, n_total, n_late, percent_late, n_admitedIP, percent_IP)


# Payer barplot overrall
par(mar=c(5,8,4,2))
barplot(table(ORinfo$payer), horiz = TRUE, cex.names=0.8)
title(main = "ERU patients by payor", xlab = "Number of patients")

# Payer barplot: Late patients
barplot(table(days_late$payer), horiz = TRUE, cex.names=0.8)
title(main = "ERU patients discharged late - by payor", xlab = "Number of patients")

# Table of late and admitted patients by payer
payer <- NULL
n_total <- NULL
n_late <- NULL
percent_late <- NULL
n_admitedIP <- NULL
percent_IP <- NULL
for(i in 1:length(levels(ORinfo$payer))){
  payer[i] <- levels(ORinfo$payer)[i]
  n_total[i] <- nrow(ORinfo[ORinfo$payer == levels(ORinfo$payer)[i],])
  n_late[i] <- nrow(ORinfo[ORinfo$payer == levels(ORinfo$payer)[i] & ORinfo$ERU_outcome == "Late",])
  percent_late[i] <- paste(round(n_late[i]/n_total[i]*100, 0), "%")
  n_admitedIP[i] <- nrow(ORinfo[ORinfo$payer == levels(ORinfo$payer)[i] & ORinfo$ERU_outcome == "IP",])
  percent_IP[i] <- paste(round(n_admitedIP[i]/n_total[i]*100, 0), "%")
}
late_prop_payer <- data.frame(payer, n_total, n_late, percent_late, n_admitedIP, percent_IP)

par(mar = rep(3, 4))
par(mfrow=c(3,1)) 
# OR day of week for on time patients
paste(round(prop.table(table(days_ontime$weekday))*100,2), "%")
par(las=1)
barplot(prop.table(table(days_ontime$weekday)))
title(main = "On-time discharges")

# OR day of week for late patients
paste(round(prop.table(table(days_late$weekday))*100,2), "%")
barplot(prop.table(table(days_late$weekday)))
title(main = "Late discharges")

# OR day of week for patients who were admitted inpatient
paste(round(prop.table(table(days_IP$weekday))*100,2), "%")
barplot(prop.table(table(days_IP$weekday)))
title(main = "Admitted IP")

# Table of late and admitted patients by day of week
day_of_week <- NULL
n_total <- NULL
n_ontime <- NULL
percent_ontime <- NULL
n_late <- NULL
percent_late <- NULL
n_admitedIP <- NULL
percent_IP <- NULL
for(i in 1:length(levels(ORinfo$weekday))){
  day_of_week[i] <- levels(ORinfo$weekday)[i]
  n_total[i] <- nrow(ORinfo[ORinfo$weekday == levels(ORinfo$weekday)[i],])
  n_ontime[i] <- nrow(ORinfo[ORinfo$weekday == levels(ORinfo$weekday)[i] & ORinfo$ERU_outcome == "On_time",])
  percent_ontime[i] <- paste(round(n_ontime[i]/n_total[i]*100, 0), "%")
  n_late[i] <- nrow(ORinfo[ORinfo$weekday == levels(ORinfo$weekday)[i] & ORinfo$ERU_outcome == "Late",])
  percent_late[i] <- paste(round(n_late[i]/n_total[i]*100, 0), "%")
  n_admitedIP[i] <- nrow(ORinfo[ORinfo$weekday == levels(ORinfo$weekday)[i] & ORinfo$ERU_outcome == "IP",])
  percent_IP[i] <- paste(round(n_admitedIP[i]/n_total[i]*100, 0), "%")
}
prop_weekday <- data.frame(day_of_week, n_total, n_ontime, percent_ontime, n_late, percent_late, n_admitedIP, percent_IP)

```

# Service, payor and day of week of OR split by ERU outcomes:
```{r, results='markup'}
# OR service 
arrange(late_prop_service, desc(n_total))

# Payer
arrange(late_prop_payer, desc(n_total))

# Day of week of OR
prop_weekday
```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Data cleaning
# Post-op timeline
postop_chron <- select(eru, ID:ERU_outcome, TimeoutOR_after1800, ERU_after2100, PACU_event:MDplan_after7am)
str(postop_chron)

# Transforming datatype
# For TimeoutOR_after1800
levels(postop_chron$TimeoutOR_after1800)[3] <- NA
postop_chron$TimeoutOR_after1800 <- as.character(postop_chron$TimeoutOR_after1800)
postop_chron$TimeoutOR_after1800 <- as.integer(postop_chron$TimeoutOR_after1800)
# For dc_edu_POD1
levels(postop_chron$dc_edu_POD1)[c(4, 5)] <- NA
levels(postop_chron$dc_edu_POD1)[3] <- 2
postop_chron$dc_edu_POD1 <- as.character(postop_chron$dc_edu_POD1)
postop_chron$dc_edu_POD1 <- as.integer(postop_chron$dc_edu_POD1)
# For MDplan_after7am
levels(postop_chron$MDplan_after7am)[4] <- NA
levels(postop_chron$MDplan_after7am)[3] <- 2
postop_chron$MDplan_after7am <- as.character(postop_chron$MDplan_after7am)
postop_chron$MDplan_after7am <- as.integer(postop_chron$MDplan_after7am)

#Subsetting data
timing <- select(postop_chron, ERU_outcome, TimeoutOR_after1800, ERU_after2100, dc_edu_POD1, MDplan_after7am)

# Splitting data by ERU outcome
t_ontime <- timing[timing$ERU_outcome == "On_time", 2:5]
t_late <- timing[timing$ERU_outcome == "Late", 2:5 ]
t_IP <- timing[timing$ERU_outcome == "IP", 2:5 ]
summary(t_ontime) # Confirmed has 1 missing values for TimeoutOR_after1800
summary(t_late)
summary(t_IP) # Confirmed has 1 missing values for TimeoutOR_after1800

# On time:
chron_ontime <- do.call(data.frame, list(
    proportion = paste(round(apply(t_ontime, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(t_ontime, 2, length),
    n = apply(t_ontime, 2, sum, na.rm = TRUE),
    missing_value = apply (t_ontime, 2, 
    function(x) sum(length(which(is.na(x))))
                )))
# Late:
chron_late <- do.call(data.frame, list(
    proportion = paste(round(apply(t_late, 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(t_late, 2, length),
    n = apply(t_late, 2, sum, na.rm = TRUE),
    missing_value = apply (t_late, 2, 
    function(x) sum(length(which(is.na(x))))
                )))
# Admitted IP:
chron_IP <- do.call(data.frame, list(
    proportion = paste(round(apply(t_IP[,1:2], 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(t_IP[,1:2], 2, length),
    n = apply(t_IP[,1:2], 2, sum, na.rm = TRUE),
    missing_value = apply (t_IP[,1:2], 2, 
    function(x) sum(length(which(is.na(x))))
                )))

```

## Types of possibly relevant PACU and ERU events:
```{r}
# These events have not been tabulated yet

# Types of events that happened in PACU
levels(postop_chron$PACU_event)

# TYpes of Events that happened in ERU
levels(postop_chron$ERU_event)

# Types of Multidiciplinary consults on POD1
levels(postop_chron$MDT_consult_POD1)
```

## Post-op PACU, ERU and discharge events Timeline:
```{r}
# For patients who were discharged on time:
chron_ontime

# For patients who were discharged late:
chron_late

# For patients who were admitted inpatient:
chron_IP
```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Data cleaning
# Reasons
reasons <- select(eru, ID:ERU_outcome, Prim_reason, other_reasons, doc_ride_issue, last_minute_foley)
# str(reasons)

# Transform data type for doc_ride_issue and last_minute_foley
levels(reasons$doc_ride_issue)[3] <- NA
reasons$doc_ride_issue <- as.character(reasons$doc_ride_issue)
reasons$doc_ride_issue <- as.integer(reasons$doc_ride_issue)
levels(reasons$last_minute_foley)[3] <- NA
reasons$last_minute_foley <- as.character(reasons$last_minute_foley)
reasons$last_minute_foley <- as.integer(reasons$last_minute_foley)

# Split by ERU outcome
r_ontime <- reasons[reasons$ERU_outcome == "On_time",]
r_late <- reasons[reasons$ERU_outcome == "Late",]
r_IP <- reasons[reasons$ERU_outcome == "IP",]

# On time:
rf_ontime <- do.call(data.frame, list(
    proportion = paste(round(apply(r_ontime[, 5:6], 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(r_ontime[, 5:6], 2, length),
    n = apply(r_ontime[, 5:6], 2, sum, na.rm = TRUE),
    missing_value = apply (r_ontime[, 5:6], 2, 
    function(x) sum(length(which(is.na(x))))
                )))
# Late:
rf_late <- do.call(data.frame, list(
    proportion = paste(round(apply(r_late[, 5:6], 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(r_late[, 5:6], 2, length),
    n = apply(r_late[, 5:6], 2, sum, na.rm = TRUE),
    missing_value = apply (r_late[, 5:6], 2, 
    function(x) sum(length(which(is.na(x))))
                )))

```

## Brief glance at the various reasons for late discharges or admission:
```{r}
# Primary reasons late discharges or admission:
levels(reasons$Prim_reason)
# Other reasons late discharges or admission:
levels(reasons$other_reasons)

# Create a vector of reasons to create word cloud - giving primary reason twice the weight:

words <- c(as.character(reasons$Prim_reason), as.character(reasons$Prim_reason), as.character(reasons$other_reasons))
themes <- c("bleeding", "BSL", "cardiac_issue", "cardio_consult", "case_coordinator", "corneal_abrasion_postop", "CXR", "dc_edu", "drain", "failed_PT", "fever", "foley_edu", "hematoma", "hematuria", "hemodialysis", "interpreter", "intraop_blood_loss", "unclear", "iv_med", "lab_result", "late_TOV", "low_Ca", "lethargy", "low_UO", "MD_plan", "VNA", "PT", "SW", "med_4_itch", "med_due_8am", "migraine", "mobility", "nausea", "nerve_palsy_postop", "O2sat", "OR_postponed", "oral_intake", "pain_control", "patient_request", "perforation_postop", "refused_dc", "rehab_dc", "returned_OR", "ride_issue", "social_issue", "poor_note", "very_sedated_postop", "voiding_issue", "wound", "background_comorbidities", "BP", "JP_monitoring", "BSL", "opthal_consult", "pharm_rv")
count <- NULL
for (i in 1:length(themes)){
  count[i] <- length(grep(themes[i], words))
}
theme_cloud <- data.frame(cbind(count, themes))
# Export to tab delimited text file
write.table(theme_cloud, "late_theme_cloud.txt", sep="\t", row.names=FALSE, quote=FALSE)
```

## Proportion of ride and last-minute foley issues amongst on_time and late discharges:
```{r}
# Ride issue is defined as documented waiting for ride as part of reason for being late - hence none of the on time patients have 'ride issue'
# However, last minute foley is any who had a foley placed in the morning of POD1 in order to be discharged on the day

# Amongst on time discharge patients
rf_ontime

# Amongst late discharge patients
rf_late

```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results='hide'}
# Data cleaning
# ERU discharge quality indicators
quality_dc <- select(eru, ID:ERU_outcome, last_minute_foley:patient_not_ready4dc)
quality_dc <- filter(quality_dc, ERU_outcome != "IP")
str(quality_dc)

# On time ERU discharges
summary(quality_dc[quality_dc$ERU_outcome == "On_time", 3:5])
# Late ERU discharges
summary(quality_dc[quality_dc$ERU_outcome == "Late", 3:5])

# Transforming data
levels(quality_dc$last_minute_foley)[3] <- NA
quality_dc$last_minute_foley <- as.integer(quality_dc$last_minute_foley)
levels(quality_dc$Re.admission.within.3.days)[c(3,4)] <- NA
quality_dc$Re.admission.within.3.days <- as.integer(quality_dc$Re.admission.within.3.days)
levels(quality_dc$patient_not_ready4dc)[c(3,4)] <- NA
quality_dc$patient_not_ready4dc <- as.integer(quality_dc$patient_not_ready4dc)

# Splitting by ERU outcome
q_ontime <- quality_dc[quality_dc$ERU_outcome == "On_time",]
q_late <- quality_dc[quality_dc$ERU_outcome == "Late",]

# On time:
Q_ontime <- do.call(data.frame, list(
    proportion = paste(round(apply(q_ontime[, 3:5], 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(q_ontime[, 3:5], 2, length),
    n = apply(q_ontime[, 3:5], 2, sum, na.rm = TRUE),
    missing_value = apply (q_ontime[, 3:5], 2, 
    function(x) sum(length(which(is.na(x))))
                )))
# Late:
Q_late <- do.call(data.frame, list(
    proportion = paste(round(apply(q_late[, 3:5], 2, mean, na.rm = TRUE)*100,2), " %"),
    n_length = apply(q_late[, 3:5], 2, length),
    n = apply(q_late[, 3:5], 2, sum, na.rm = TRUE),
    missing_value = apply (q_late[, 3:5], 2, 
    function(x) sum(length(which(is.na(x))))
                )))
```

## Indicators of likely poor ERU discharges for On-time and late discharges:
```{r}
# Patients admitted have been excluded
# Discharge quality indicators are:
# 1. Last minute foley
# 2. Re-admission to Partner's network shown on LMR system within 3 days 
# 3. Patient stating 'not ready' for discharge on the 2 day nursing phone follow-up questionnaire

# On time discharges
Q_ontime

# Late discharges
Q_late
```

# Urology patients analysis
```{r}
eru %>% group_by(service) %>% tally
uro <- eru[eru$service == "URO", c("ERU_outcome", "Age", "Gender", "ASA_34", "service","Prim_reason", "other_reasons", "last_minute_foley")]

summary(uro$ERU_outcome)
prop.table(table(uro$ERU_outcome))
summary(uro$Age)
summary(uro$Gender)
summary(uro$ASA_34)
sort(prop.table(table(uro$Prim_reason)), decreasing = TRUE)
sort(prop.table(table(uro$other_reasons)), decreasing = TRUE)
foley_uro <- uro[uro$ERU_outcome != "IP", c("ERU_outcome", "last_minute_foley")]
summary(foley_uro$last_minute_foley)
foley_uro
```

# Gynecology patients analysis
```{r}
eru %>% group_by(service) %>% tally
gyne <- eru[eru$service == "GYN", c("ERU_outcome", "Age", "Gender", "ASA_34", "service","Prim_reason", "other_reasons", "last_minute_foley")]

nrow(gyne)
summary(gyne$ERU_outcome)
paste(round(prop.table(table(gyne$ERU_outcome))*100,2), "%")
summary(gyne$Age)
summary(gyne$Gender)
summary(gyne$ASA_34)
sort(prop.table(table(gyne$Prim_reason)), decreasing = TRUE)
sort(prop.table(table(gyne$other_reasons)), decreasing = TRUE)
foley_gyn <- gyne[gyne$ERU_outcome != "IP", c("ERU_outcome", "last_minute_foley")]
summary(foley_gyn$last_minute_foley)
foley_gyn
```

# Plastic patients analysis
```{r}
eru %>% group_by(service) %>% tally
plas <- eru[eru$service == "PLA", c("ERU_outcome", "Age", "Gender", "ASA_34", "service","Prim_reason", "other_reasons", "last_minute_foley")]

nrow(plas)
summary(plas$ERU_outcome)
paste(round(prop.table(table(plas$ERU_outcome))*100,2), "%")
summary(plas$Age)
summary(plas$Gender)
summary(plas$ASA_34)
sort(prop.table(table(plas$Prim_reason)), decreasing = TRUE)
sort(prop.table(table(plas$other_reasons)), decreasing = TRUE)
foley_pla <- plas[plas$ERU_outcome != "IP", c("ERU_outcome", "last_minute_foley")]
summary(foley_pla$last_minute_foley)
foley_pla
```

# GGI patients analysis
```{r}
eru %>% group_by(service) %>% tally
ggi <- eru[eru$service == "GGI", c("ERU_outcome", "Age", "Gender", "ASA_34", "service","Prim_reason", "other_reasons", "last_minute_foley")]

nrow(ggi)
summary(ggi$ERU_outcome)
paste(round(prop.table(table(ggi$ERU_outcome))*100,2), "%")
summary(ggi$Age)
summary(ggi$Gender)
summary(ggi$ASA_34)
sort(prop.table(table(ggi$Prim_reason)), decreasing = TRUE)
sort(prop.table(table(ggi$other_reasons)), decreasing = TRUE)
foley_ggi <- ggi[ggi$ERU_outcome != "IP", c("ERU_outcome", "last_minute_foley")]
summary(foley_ggi$last_minute_foley)
foley_ggi
```

# ANE patients analysis
```{r}
eru %>% group_by(service) %>% tally
ane <- eru[eru$service == "ANE", c("ERU_outcome", "Age", "Gender", "ASA_34", "service","Prim_reason", "other_reasons", "last_minute_foley")]

nrow(ane)
summary(ane$ERU_outcome)
paste(round(prop.table(table(ane$ERU_outcome))*100,2), "%")
summary(ane$Age)
summary(ane$Gender)
summary(ane$ASA_34)
sort(prop.table(table(ane$Prim_reason)), decreasing = TRUE)
sort(prop.table(table(ane$other_reasons)), decreasing = TRUE)
foley_ane <- ane[ane$ERU_outcome != "IP", c("ERU_outcome", "last_minute_foley")]
summary(foley_ane$last_minute_foley)
foley_ane
```

# Further work to be done include:
1. Teasing out events in PACU and ERU
2. Teasing out reasons for late discharge and admission
3. Teasing out service and payor info into ERU outcome categories
4. Co-morbidity information still needs cleaning at Excel stage
5. Further exploratory analysis and logistical regression model fitting

